A Note on Convergence Diagnostics for Multiple Imputation Algorithms

### 2019-12-12 Simulation/Writeup/NoteConvergence.Rnw
definition rhat

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{A Note on Convergence Diagnostics for Multiple Imputation using Chained Equations (MICE)}
\Plaintitle{A Note on Convergence Diagnostics for Multiple Imputation using Chained Equations (MICE)}
\Shorttitle{Convergence Diagnostics for MICE}

%% - \Abstract{} almost as usual
\Abstract{
The ubiquitous problem of missing information may be circumvented by implementing the framework of multiple imputation (MI). With MI, missing data points are `filled in' (i.e. imputed) several times, and statistical inference is performed on each imputed data set. The validity of inferences may be threatened by non-convergence of this iterative algorithmic procedure, but diagnostics to evaluate algorithmic convergence have not been implemented in MI software. This note investigates the performance of two convergence diagnostics ($\widehat{R}$ \citep{gelm92}, auto-correlation \citep{scha97}) through model-based simulation. To investigate convergence of the algorithm, the number of iterations is varied between one and one-hundred. Each of these 100 simulation conditions is replicated 1000 times. The convergence diagnostics are evaluated against three recommended simulation diagnostics (bias, confidence interval width, and coverage rate \citep{buur18}, since no benchmark diagnostic is available. This simulation study shows that three iterations may be sufficient to obtain ... \textbf{This Research Report contains a simulation study that serves as the basis of the technical paper that will be submitted for publication in \emph{Journal of Statistical Software}. I aim to publish a pre-print of this research report on \href{https://arxiv.org/}{arXiv}. That way, I can refer to the simulation study described here, without `bulking up' the technical paper that I want to submit for publication in JSS. Another option would be to attach this note as an online appendix to the technical paper on ShinyMICE.}  
% This Research Report contains a simulation study that serves as the basis of the technical paper that will be submitted for publication in \emph{Journal of Statistical Software}. I have chosen to use the first format from the Research Report guidelines: ``\emph{It is written as a (mini) thesis, with an introduction, methods section, some results (i.e., preliminary analyses, or pilot simulations), and a discussion of results. The length of the research report should be maximally 2500 words of text (without references list and or tables and figures). Please do not include appendices, and no more than 6 tables or figures. Table and Figure captions do not count towards the word limit. An abstract may be included, but is not necessary}''. I aim to publish a pre-print of this research report on \href{https://arxiv.org/}{ArXiv}. That way, I can refer to the simulation study described here, without 'bulking up' the technical paper that I want to submit for publication in JSS. Another option would be to attach this note as online appendix to the technical paper on ShinyMICE.
  }

@@

The goal of this research project is to develop novel methodology and guidelines for evaluating MI methods. These evaluation measures and guidelines will subsequently be implemented in an interactive evaluation framework for multiple imputation to aid applied researchers in drawing valid inference from incomplete datasets. This note provides the theoretical foundation towards the diagnostic evaluation of the convergence of MI algorithms. % for diagnostics for evalua one of these evaluation measures that is vital in evaluating MI procedures: a diagnostic to assess convergence of the MI algorithm. 
For reasons of brevity, this note only focuses ion the MI algo implemented in mice. %The implementation limits itself to the MI algorithm 
%More specifically, this note focuses on the MI algorithm that is implemented in the \proglang{R} package \pkg{mice}: `Multiple Imputation using Chained Equations' (MICE) \textbf{[can be removed, can be found in the mice ref]}. 
To this end we will evaluate how convergence of the imputation algo can be diagnosed.

@@

The convergence properties of MICE are investigated through model-based simulation in \proglang{R} \citep{R}. The results of this simulation study are guidelines for assessing convergence of the MI algorithm. These guidelines will be implemented in an interactive evaluation tool for \pkg{mice}, `ShinyMICE', which is currently under development. All programming code used in this note is available from \href{https://github.com/gerkovink/shinyMice/simulation}{github.com/gerkovink/ShinyMICE/simulation}. 
%I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence. Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{auto-correlation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.

@@

\begin{figure}
\label{fig:steps}
\centering
	\large{\begin{tikzpicture}
	\tikzstyle{main}=[circle, minimum size = 10mm, thick, draw =black!80, node distance = 20mm]
\tikzstyle{connect}=[-latex, thick]
\node[main, fill = white!100] (data) [label=left:Incomplete data] { };
\node[main] (mids) [right=of data] { };
\node[main] (mids2) [above=10mm of mids] { };
\node[main] (mids3) [below=10mm of mids,label=below:Imputed data] { };
\node[main] (mira) [right=of mids] {};
\node[main] (mira2) [above=10mm of mira] { };
\node[main] (mira3) [below=10mm of mira,label=below:Analysis results] { };
\node[main, fill = white!100] (mipo) [right=of mira,label=right:Pooled result] { };
\path (data) edge [connect] (mids)
      (data) edge [connect] (mids2)
      (data) edge [connect] (mids3)
      (mids) edge [connect] (mira)
      (mids2) edge [connect] (mira2)
      (mids3) edge [connect] (mira3)
		  (mira) edge [connect] (mipo)
		  (mira2) edge [connect] (mipo)
		  (mira3) edge [connect] (mipo);
\end{tikzpicture}}
\caption{Scheme of the main steps in multiple imputation ($m = 3$). Adapted from \citep[\S~1.4.1]{buur18}.}
\end{figure}

@@

This note focuses on the algorithmic properties of the imputation step. The algorithm employed within this step has an iterative nature. That is, before drawing $m$ imputed values for each missing data point in $Y_{mis}$, a `chain' of potential values is sampled. Each of the $m$ chains starts with an initial value, drawn randomly from $Y_{obs}$. The chains are terminated after a predefined number of iterations. Only the ultimate sample that a chain lands on is imputed and subsequently used in the analysis and pooling steps. The collection of samples across iterations (between the initial value and the imputed value) for each of the $m$ imputations will be referred to as an `imputation chain'. The number of iterations will be denoted with $t$ \textbf{[NOG TOEPASSEN]}.

@@

The multiple imputation algorithm `MICE' is a special case of Markov chain Monte Carlo (MCMC) methods--the general framework of iterative algorithmic procedures. While methodology for diagnosing convergence of multiple imputation algorithms has not been systematically studied \citep{buur18}, several convergence diagnostics have been proposed for other types of MCMC methods (e.g., Gibbs samplers), or the framework in general.
% For the framework of itrIt is not known whether convergence diagnostics that are available for the general framework that MICE is a special case of are appropriate for the evaluation of MI data. This is the framework of Markoc chain Monte Carlo %Yet, reference manuals of promiment MI software do mention the importance of evaluating algorithmic convergence (e.g., \cite[p.XYZ]{mice}, \textbf{XYZ, XYZ}). 
This section provides a brief review of: 1) how convergence is defined in the general context of iterative algorithmic procedures; 2) what diagnostic tools are available; and 3) which of these may apply to MI algorithms.

Loosely speaking, the aim of MCMC algorithms (e.g., Gibbs samplers, or MICE) is to sample values from an unknown target distribution (e.g., the posterior distribution of a parameter in Bayesian inference, or a distribution of missing values, respectively). Convergence is reached once the algorithm samples exclusively from the target distribution \citep{cowl96}. Since this target distribution is unknown by definition, convergence can only be monitored by evaluating signs of non-convergence \citep{hoff09}. Therefore, in practice, the definition of convergence consists of two components: `mixing' of chains together, and `stationarity' of individual chains \citep[p.~284]{gelm13}. The mixing component is satisfied when chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm. Stationarity refers to the absence of trending within chains. \textbf{Figure XYZ shows a pathological case of non-convergence concerning both components.}  
%The aim of MCMC procedures is to sample from an unknown (theoretical) distribution. The stability of the solution denpends on two components of convergence: should have converged, two components should be satisfied.

@@

The two components of convergence may be evaluated with convergence diagnostics. Diagnostics that target the stationarity component are auto-correlation \citep{scha97, gelm13}, numeric standard error \citep[which is commonly called `MC error']{gewe92}, and Raftery and Lewis's \citeyear{raft91} procedure to determine the increase in the necessary number of iterations due to dependence within chains. The mixing component of convergence, on the other hand, is primarily assessed with the `Gelman-Rubin statistic'--potential scale reduction factor $\widehat{R}$ \citep{gelm92}. An adapted version of $\widehat{R}$ has been proposed by \cite{veht19}, which may also assess the stationarity component of convergence. Therefore, the focus of this study is mainly on $\widehat{R}$. Stationarity will also be evaluated with auto-correlation as diagnostic, as recommended by \cite[p.~898]{cowl96}. Other convergence diagnostics are outside the scope of this study. \footnotemark 
\footnotetext{All of these methods evaluate the convergence of univariate scalar summaries (e.g., chain means or variances). These convergence diagnostics cannot diagnose convergence of multivariable statistics (i.e., relations between scalar summaries). \cite{buur18} proposed to implement multivariable evaluation by means of eigenvalue decomposition \cite{mack03}. This method is outside of the scope of the current study.}%  he eigenvector decomposition method proposed by McKay \citep{XYZ}.} These four methods, however, are outside of the scope of this study. Here, the focus is mainly on $\widehat{R}$ and auto-correlation.

@@

It is not known whether any convergence diagnostic is appropriate to diagnose convergence of the MICE algorithm. As recommended by \cite[\S~6.5.2]{buur18}, the current practice is to visually inspect imputation chains for signs of non-mixing and non-stationarity. Not only is visually assessing convergence challenging to the untrained eye--it might also be futile. That is, there is no scientific consensus on the convergence properties of multiple imputation algorithms \citep{taka17}, and in some situations, the default techniques in \pkg{mice} might not converge to stable target distributions at all \citep{murr18}. %Convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.  Moreover, 
%A fundamental assumption of MICE is convergence of the algorithm. MICE is a type of Markov chain Monte Carlo (MCMC) algorithm. In general, the validity of inference resulting from MCMC algorithms is threatened by non-convergence. Hence we use convergence diagnostics to flag non-convergence. But it is not known whether conventional convergence diagnostics for MCMC methods work on MI data.
Therefore, the need for convergence diagnostics is extra pressing in the context of the MICE algorithm.

@@

\textbf{Explain the measures that I'm investigating here. R hat is ... Auto-correlation is... How are they computed? What do they say? How are they computed on MI data? Looking at imputation chains. 
Explain that there is no baseline to compare the measures with. The only current convergence check is to visually inspect trace plots. (But in the trace-plots convergence seems immediate).}

The potential scale reduction factor $\widehat{R}$ tells us by how much the variance of an estimate could be shrunken down if the chains were to be infinitely long \cite{gelm92}. That then informs us about how dependent the chains are on the starting values. If there is no dependence on the initial values anymore, the chains have converged (in the mixing sense of the word). $\widehat{R}$ is then equal to one. The conventionally acceptable threshold was $\widehat{R} < 1.2$. More recently, \cite{veht19} proposed a more stringent threshold: $\widehat{R} < 1.01$.

%We can apply $\widehat{R}$ to the mean (or to the first two moments) of the variables of interest. 
$\widehat{R}$ is defined by \cite{veht19} as follows: % The scalar summary of interest in this study is the average of the imputed values per variable. %As recommended by \cite{veht19} $\widehat{R}$ is computed as .... to be able to detect ... in the tails of the distribution. \cite[p.~5]{veth19}

\textbf{REPHRASE: In the equations below, $N$ is the number of draws per chain, $M$ is the number of chains, and $S = MN$ is the total number of draws from all chains. For each scalar summary of interest $\theta$, we compute $B$ and $W$, the between- and within-chain variances:}

\begin{align*}
``B&=\frac{N}{M-1} \sum_{m=1}^{M}\left(\bar{\theta}^{(\cdot m)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot m)}=\frac{1}{N} \sum_{n=1}^{N} \theta^{(n m)}, \quad \bar{\theta}^{(\cdot \cdot)}=\frac{1}{M} \sum_{m=1}^{M} \bar{\theta}^{(\cdot m)} \\
W&=\frac{1}{M} \sum_{m=1}^{M} s_{j}^{2},  \text { where } s_{m}^{2}=\frac{1}{N-1} \sum_{n=1}^{N}\left(\theta^{(n m)}-\bar{\theta}^{(\cdot m)}\right)^{2}". \text{ \cite[p.~5]{veht19}} 
\end{align*}

The proportion of within chain variance $W$ against the weighted average of $B$ and $W$, $\widehat{\operatorname{var}}^{+}$ tells us how much variance of $\theta$ could be shrunken down if $N\to\infty$. $\widehat{\operatorname{var}}^{+}$ is computed as follows: %over-estimates the marginal posterior variance of the estimand, $\theta$.

\begin{equation*}
\widehat{\operatorname{var}}^{+}(\theta | y)=\frac{N-1}{N} W+\frac{1}{N} B.
\end{equation*}

The potential scale reduction factor $\widehat{R}$ is obtained as:

\begin{equation*}
\widehat{R}=\sqrt{\frac{\widehat{\operatorname{var}}^{+}(\theta | y)}{W}}.
\end{equation*}

High $\widehat{R}$ would indicate .... low $\widehat{R}$ values indicate ....

The second component of convergence stationarity requires chains to be stable across iterations, or non-recurring. Non-recurrence can be evaluated with auto-correlation. Auto-correlation shows how dependent subsequent draws of an imputation chain are on the previous value. If there is a lot of dependence, draws at e.g. iteration five are significantly correlated with the value of the first draw. A high auto-correlation indicates dependence within chains. The magnitude of the AC can be interpreted qualitatively or quantitatively. Quantitative evaluation of the AC entails comparing the observed ACs to the critical values of a two-tailed 95\% confidence interval, divided by the square root of the number of iterations. %. critical value corresponding to the $z-$~value standard normal distribution with \textbf{XYZ} degrees of freedom.
%  AC_lim <- qnorm(.975) / sqrt(maxit)

@@

The aim is to evaluate whether the imputation procedure has converged. The primary research interest is in determining whether $\widehat{R}$ is an appropriate convergence diagnostic, and if so, which level of stringency suits MI data. \textbf{Or if there is a discrepancy between the visual and numeric inspection.}

@@

This paper is written by the sole author (Hanne Oberman, BSc.), with guidance from Master thesis supervisors prof. dr. Stef van Buuren, and dr. Gerko Vink.

>>>>

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{A Note on Convergence Diagnostics for Multiple Imputation Algorithms}
\Plaintitle{A Note on Convergence Diagnostics for Multiple Imputation Algorithms}
\Shorttitle{Convergence Diagnostics for MI}

%% - \Abstract{} almost as usual
\Abstract{
The ubiquitous problem of missing information may be circumvented by implementing the framework of multiple imputation (MI). With MI, missing data points are `filled in' (i.e. imputed) several times, and statistical inference is performed on each imputed data set. The validity of inferences may be threatened by non-convergence of this iterative algorithmic procedure, but diagnostics to evaluate algorithmic convergence have not been implemented in MI software. This note investigates the performance of two convergence diagnostics (\citep[$\widehat{R}$,][]{gelm92}; \citep[auto-correlation,][]{scha97}) through model-based simulation. To investigate convergence of the algorithm, the number of iterations is varied between one and one-hundred. Each of these 100 simulation conditions is replicated 1000 times. The convergence diagnostics are evaluated against three recommended simulation diagnostics (bias, confidence interval width, and coverage rate \citep{buur18}, since no benchmark diagnostic is available. This simulation study shows that three iterations may be sufficient to obtain ... \textbf{This Research Report contains a simulation study that serves as the basis of the technical paper that will be submitted for publication in \emph{Journal of Statistical Software}. I aim to publish a pre-print of this research report on \href{https://arxiv.org/}{arXiv}. That way, I can refer to the simulation study described here, without `bulking up' the technical paper that I want to submit for publication in JSS. Another option would be to attach this note as an online appendix to the technical paper on ShinyMICE.}  
% This Research Report contains a simulation study that serves as the basis of the technical paper that will be submitted for publication in \emph{Journal of Statistical Software}. I have chosen to use the first format from the Research Report guidelines: ``\emph{It is written as a (mini) thesis, with an introduction, methods section, some results (i.e., preliminary analyses, or pilot simulations), and a discussion of results. The length of the research report should be maximally 2500 words of text (without references list and or tables and figures). Please do not include appendices, and no more than 6 tables or figures. Table and Figure captions do not count towards the word limit. An abstract may be included, but is not necessary}''. I aim to publish a pre-print of this research report on \href{https://arxiv.org/}{ArXiv}. That way, I can refer to the simulation study described here, without 'bulking up' the technical paper that I want to submit for publication in JSS. Another option would be to attach this note as online appendix to the technical paper on ShinyMICE.
  }

@@

The goal of this research project is to develop novel methodology and guidelines for evaluating MI methods. These evaluation measures and guidelines will subsequently be implemented in an interactive evaluation framework for multiple imputation to aid applied researchers in drawing valid inference from incomplete datasets. This note provides the theoretical foundation towards the diagnostic evaluation of the convergence of MI algorithms. % for diagnostics for evalua one of these evaluation measures that is vital in evaluating MI procedures: a diagnostic to assess convergence of the MI algorithm. 
For reasons of brevity, this note only focuses on the MI algorithm implemented in \pkg{mice} \cite{mice}. %The implementation limits itself to the MI algorithm 
%More specifically, this note focuses on the MI algorithm that is implemented in the \proglang{R} package \pkg{mice}: `Multiple Imputation using Chained Equations' (MICE) \textbf{[can be removed, can be found in the mice ref]}. 
To this end we will evaluate how convergence of the imputation algorithm can be diagnosed.

@@

The convergence properties of the MI algoritm in \pkg{mice} are investigated through model-based simulation. The results of this simulation study are guidelines for assessing convergence of MI algorithms. These guidelines will be implemented in an interactive evaluation tool for \pkg{mice}, `ShinyMICE', which is currently under development. All programming code used in this note is available from \href{https://github.com/gerkovink/shinyMice/simulation}{github.com/gerkovink/ShinyMICE/simulation}. 
%I will replicate Lacerda et al.'s simulation study on $\widehat{R}$ \citep{lace07}, and develop novel guidelines for assessing convergence. Ideally, I will integrate several diagnostics (e.g., $\widehat{R}$, \emph{auto-correlation}, and \emph{simulation error}) into a single summary indicator to flag non-convergence.

@@

\begin{figure}
\label{fig:steps}
\centering
	\large{\begin{tikzpicture}
	\tikzstyle{main}=[circle, minimum size = 10mm, thick, draw =black!80, node distance = 20mm]
\tikzstyle{connect}=[-latex, thick]
\node[main, fill = white!100] (data) [label=left:Incomplete data] { };
\node[main] (mids) [right=of data] { };
\node[main] (mids2) [above=10mm of mids] { };
\node[main] (mids3) [below=10mm of mids,label=below:Imputed data] { };
\node[main] (mira) [right=of mids] {};
\node[main] (mira2) [above=10mm of mira] { };
\node[main] (mira3) [below=10mm of mira,label=below:Analysis results] { };
\node[main, fill = white!100] (mipo) [right=of mira,label=right:Pooled result] { };
\path (data) edge [connect] (mids)
      (data) edge [connect] (mids2)
      (data) edge [connect] (mids3)
      (mids) edge [connect] (mira)
      (mids2) edge [connect] (mira2)
      (mids3) edge [connect] (mira3)
		  (mira) edge [connect] (mipo)
		  (mira2) edge [connect] (mipo)
		  (mira3) edge [connect] (mipo);
\end{tikzpicture}}
\caption{Scheme of the main steps in multiple imputation ($m = 3$). Adapted from \cite[\S~1.4.1]{buur18}.}
\end{figure}

@@

This note focuses on the algorithmic properties of the imputation step. The algorithm employed within this step has an iterative nature. That is, before drawing $m$ imputed values for each missing data point in $Y_{mis}$, a `chain' of potential values is sampled. Each of the $m$ chains starts with an initial value, drawn randomly from $Y_{obs}$. The chains are terminated after a predefined number of iterations. Only the ultimate sample that a chain lands on is imputed and subsequently used in the analysis and pooling steps. The collection of samples across iterations (between the initial value and the imputed value) for each of the $m$ imputations will be referred to as an `imputation chain'. The number of iterations will be denoted with $t$.

@@

The multiple imputation algorithm in \pkg{mice} is a special case of Markov chain Monte Carlo (MCMC) methods--a framework of iterative algorithmic procedures. While methodology for diagnosing convergence of MI algorithms has not been systematically studied \citep{buur18}, several convergence diagnostics have been proposed for other types of MCMC methods (e.g., Gibbs samplers) and the framework in general.
% For the framework of itrIt is not known whether convergence diagnostics that are available for the general framework that MICE is a special case of are appropriate for the evaluation of MI data. This is the framework of Markoc chain Monte Carlo %Yet, reference manuals of promiment MI software do mention the importance of evaluating algorithmic convergence (e.g., \cite[p.XYZ]{mice}, \textbf{XYZ, XYZ}). 
This section provides a brief review of: 1) how convergence is defined in the general context of iterative algorithmic procedures; 2) what diagnostic tools are available; and 3) which of these may apply to MI algorithms.

Loosely speaking, the aim of MCMC algorithms (e.g., Gibbs samplers, or the MI algorithm in \pkg{mice}) is to sample values from an unknown target distribution (e.g., the posterior distribution of a parameter in Bayesian inference, or a distribution of missing values, respectively). Convergence is reached once the algorithm samples exclusively from the target distribution \citep{cowl96}. Since this target distribution is unknown by definition, convergence can only be monitored by evaluating signs of non-convergence \citep{hoff09}. In practice, the definition of convergence consists of two components: `mixing' of chains together, and `stationarity' of individual chains \citep[p.~284]{gelm13}. The mixing component is satisfied when chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm. Stationarity refers to the absence of trending across iterations within chains. %\textbf{Figure XYZ shows a pathological case of non-convergence concerning both components.}  
%The aim of MCMC procedures is to sample from an unknown (theoretical) distribution. The stability of the solution denpends on two components of convergence: should have converged, two components should be satisfied.

@@

Most convergence diagnostics target either of the two components. The stationarity component of convergence may be evaluated with auto-correlation \citep{scha97, gelm13}, numeric standard error \citep[which is commonly called `MC error']{gewe92}, or Raftery and Lewis's (\citeyear{raft91}) procedure to determine how many iterations are necessary to deminish the effect of trending within chains. The mixing component can be assessed with the potential scale reduction factor $\widehat{R}$ \citep[widely referred to as `the Gelman-Rubin statistic']{gelm92}. With an adapted version of $\widehat{R}$, proposed by \cite{veht19}, we may also evaluate the stationarity component of convergence. This would make $\widehat{R}$ a general convergence diagnostic. The application of $\widehat{R}$ to assess stationarity has not been thourougly investigated. Therefore, this study employs both $\widehat{R}$ and auto-correlation to investigate algorithmic convergence, as recommended by \cite[p.~898]{cowl96}.
%Therefore, the focus of this study is mainly on $\widehat{R}$. Stationarity will also be evaluated with auto-correlation as diagnostic, as recommended by \cite[p.~898]{cowl96}. 
Other convergence diagnostics are outside the scope of this study.\footnotemark 
\footnotetext{All of these methods evaluate the convergence of univariate scalar summaries (e.g., chain means or variances). These convergence diagnostics cannot diagnose convergence of multivariable statistics (i.e., relations between scalar summaries). \cite{buur18} proposed to implement multivariable evaluation by means of eigenvalue decomposition \cite{mack03}. This method is outside of the scope of the current study.}%  he eigenvector decomposition method proposed by McKay \citep{XYZ}.} These four methods, however, are outside of the scope of this study. Here, the focus is mainly on $\widehat{R}$ and auto-correlation.

@@

It is not known whether any convergence diagnostic is appropriate to diagnose convergence of \pkg{mice} algorithms. As by \cite[\S~6.5.2]{buur18}, the current practice is to visually inspect imputation chains for signs of non-mixing and non-stationarity. Visually assessing convergence, however, may be challenging to the untrained eye. \textbf{[This is problematic because]} There is no scientific consensus on the convergence properties of multiple imputation algorithms \citep{taka17}, and in some situations, the default techniques in \pkg{mice} might not converge to stable target distributions at all \citep{murr18}. %Convergence diagnostics for MI methods have not been systematically studied \citep{buur18}.  Moreover, 
%A fundamental assumption of MICE is convergence of the algorithm. MICE is a type of Markov chain Monte Carlo (MCMC) algorithm. In general, the validity of inference resulting from MCMC algorithms is threatened by non-convergence. Hence we use convergence diagnostics to flag non-convergence. But it is not known whether conventional convergence diagnostics for MCMC methods work on MI data.
Therefore, convergence of the \pkg{mice} algorithms should be monitored carefully. %the need for convergence diagnostics is extra pressing in the context of the MICE algorithm.

@@

The convergence diagnostics under consideration are $\widehat{R}$ and auto-correlation. Perfect convergence is reached when the variance between imputation chains is equivalent to the variance within chains ($\widehat{R} = 1$), and there is no dependency between subsequent samples within imputation chains (auto-correlation $= 0$). 
%
%%%%%%%%%%%%%%%%%
%As potential scale reduction factor, $\widehat{R}$ tells us by how much the variance of an estimate could be shrunken down if the chains were to be infinitely long \citep[$t \to \infty$;][]{gelm92}. If there is no added value compared to infinitely many iterations, the chains have converged. 
%%%%%%%%%%%%%%%%%%
%
%$\textbf{Explain the measures that I'm investigating here. R hat is ... Auto-correlation is... How are they computed? What do they say? How are they computed on MI data? Looking at imputation chains. 
%Explain that there is no baseline to compare the measures with. The only current convergence check is to visually inspect trace plots. (But in the trace-plots convergence seems immediate).}
%
%The potential scale reduction factor $\widehat{R}$ tells us by how much the variance of an estimate could be shrunken down if the chains were to be infinitely long \cite{gelm92}. That then informs us about how dependent the chains are on the starting values. If there is no dependence on the initial values anymore, the chains have converged (in the mixing sense of the word). 
%$\widehat{R}$ is then equal to one. 
%
%We can apply $\widehat{R}$ to the mean (or to the first two moments) of the variables of interest. 
To define $\widehat{R}$, we follow notation by \citep[p.~5]{veht19}. % The scalar summary of interest in this study is the average of the imputed values per variable. %As recommended by \cite{veht19} $\widehat{R}$ is computed as .... to be able to detect ... in the tails of the distribution. \cite[p.~5]{veth19}
Let $M$ be the number of chains, $N$ the number of iterations per chain, and $\theta$ the scalar summary of interest (e.g., chain mean or chain variance).
We compute the variance within each chain of $\theta$ values ($W$) and between the $m$ chains ($B$) as follows:
%\textbf{REPHRASE: In the equations below, $N$ is the number of draws per chain, $M$ is the number of chains, and $S = MN$ is the total number of draws from all chains. For each scalar summary of interest $\theta$, we compute $B$ and $W$, the between- and within-chain variances:}

\begin{align*}
B&=\frac{N}{M-1} \sum_{m=1}^{M}\left(\bar{\theta}^{(\cdot m)}-\bar{\theta}^{(\cdot \cdot)}\right)^{2}, \text { where } \bar{\theta}^{(\cdot m)}=\frac{1}{N} \sum_{n=1}^{N} \theta^{(n m)}, \quad \bar{\theta}^{(\cdot \cdot)}=\frac{1}{M} \sum_{m=1}^{M} \bar{\theta}^{(\cdot m)} \\
W&=\frac{1}{M} \sum_{m=1}^{M} s_{j}^{2},  \text { where } s_{m}^{2}=\frac{1}{N-1} \sum_{n=1}^{N}\left(\theta^{(n m)}-\bar{\theta}^{(\cdot m)}\right)^{2}. %\text{ \cite[p.~5]{veht19}} 
\end{align*}

From the between and within chain variances we compute a weighted average, $\widehat{\operatorname{var}}^{+}$, which over-estimates the total variance of $\theta$. $\widehat{R}$ is then obtained as a ratio between the over-estimated total variance and the within chain variance:

\begin{equation*}
\widehat{R}=\sqrt{\frac{\widehat{\operatorname{var}}^{+}(\theta | y)}{W}},
\text{ where } \widehat{\operatorname{var}}^{+}(\theta | y)=\frac{N-1}{N} W+\frac{1}{N} B.
\end{equation*}

%The proportion of within chain variance $W$ against the weighted average of $B$ and $W$, $\widehat{\operatorname{var}}^{+}$ tells us how much variance of $\theta$ could be shrunken down if $N\to\infty$. $\widehat{\operatorname{var}}^{+}$ is computed as follows: %over-estimates the marginal posterior variance of the estimand, $\theta$.

% \begin{equation*}
% 
% \end{equation*}

%The potential scale reduction factor $\widehat{R}$ is obtained as:

High $\widehat{R}$ would indicate .... low $\widehat{R}$ values indicate ....
The conventionally acceptable threshold was $\widehat{R} < 1.2$ \cite{gelm92}. More recently, \cite{veht19} proposed a more stringent threshold: $\widehat{R} < 1.01$.

Non-recurrence can be evaluated with auto-correlation. Auto-correlation shows how dependent subsequent draws of an imputation chain are on the previous value. If there is a lot of dependence, draws at e.g. iteration five are significantly correlated with the value of the first draw. A high auto-correlation indicates dependence within chains. The magnitude of the AC can be interpreted qualitatively or quantitatively. Quantitative evaluation of the AC entails comparing the observed ACs to the critical values of a two-tailed 95\% confidence interval, divided by the square root of the number of iterations. %. critical value corresponding to the $z-$~value standard normal distribution with \textbf{XYZ} degrees of freedom.
%  AC_lim <- qnorm(.975) / sqrt(maxit)

@@

The aim is to evaluate whether the imputation procedure has converged. The primary research interest is in determining whether $\widehat{R}$ is an appropriate convergence diagnostic, and if so, which level of stringency suits MI data. \textbf{Or if there is a discrepancy between the visual and numeric inspection. There is no baseline to compare the convergence diagnostics with. Therefore, we will only evaluate performance measures.}

@@

This note is part of a Research Master's thesis project. It was written by the sole author (Hanne Oberman, BSc.), with guidance from Master thesis supervisors prof. dr. Stef van Buuren and dr. Gerko Vink, and research seminar mentor dr. Daniela Cianci.
